My title

### 2024-12-02 paper/paper.qmd

---
title: "My title"
subtitle: "My subtitle if needed"
author: Cristina Su Lam
thanks: "Code and data are available at: https://github.com/cristinaasu/Spotify_Analysis"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
toc: true
fig_caption: yes
number-sections: true
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
# Load required libraries
library(tidyverse)
library(ggplot2)
library(hexbin)
library(arrow)
```

```{r}
# Load dataset 
analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")

```

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

@@

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variable

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-valence
#| fig-cap: Distribution of Valence
#| echo: false

ggplot(analysis_data, aes(x = Valence)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Valence", x = "Valence", y = "Count") +
  theme_minimal()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "My title"
subtitle: "My subtitle if needed"
author: Cristina Su Lam
thanks: "Code and data are available at: https://github.com/cristinaasu/Spotify_Analysis"
date: today
date-format: long
abstract: "This study investigates how musical features and contextual factors shape the emotional tone of songs, with Valence representing a track's positivity. Using a dataset of song attributes such as tempo, danceability, and acousticness, combined with population-weighted mean temperature, multiple linear regression models identified key predictors. Danceability emerged as the most significant factor, with tempo and acousticness also playing substantial roles, while temperature offered moderate explanatory power. These findings enhance our understanding of how music and context interact, with implications for personalized playlists and mood-based marketing."
toc: true
fig_caption: yes
number-sections: true
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(hexbin)
library(arrow)
library(kableExtra)
library(knitr)
library(modelsummary)
library(broom)
```

```{r}
#| include: false
#| warning: false
#| message: false
# Load dataset 
analysis_data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
mlm_model <- readRDS("../models/mlm_model.rds")
```

@@

Music has a unique power to evoke emotions, transcending cultural and linguistic barriers to connect people worldwide (Liu & Tzanetakis, 2020). Yet, understanding the emotional impact of music—captured by its valence, a measure ranging from 0 (negative or sad) to 1 (positive or happy)—remains a fascinating challenge. This study seeks to uncover the interplay between musical features such as tempo, danceability, and acousticness, and contextual factors like temperature, to better understand how these elements shape listener perceptions of positivity.

The digital age has amplified our access to music and data, creating an unprecedented opportunity for analytical exploration. Platforms like Spotify provide extensive datasets on song characteristics, enabling researchers to investigate how intrinsic musical properties and external factors converge to influence listener sentiment (Spotify for Developers, 2024). Despite this wealth of data, gaps remain in our understanding of how environmental contexts, such as weather, impact emotional responses to music. This paper aims to address this gap by examining valence as the core estimand and evaluating the significance of musical and contextual predictors.

Valence, the estimand of interest in this study, encapsulates the emotional tone of music, providing a quantitative measure of positivity. By focusing on valence, this research aims to capture the nuanced relationships between song features and listener sentiment, bridging the gap between intrinsic musical characteristics and external contextual influences.

Using a multiple linear regression (MLR) framework, we analyze data from Spotify’s top 50 Canadian charts between June and November 2024, combined with population-weighted mean temperatures from six major Canadian cities. By focusing on predictors such as danceability, acousticness, tempo, and artist-specific effects, this study identifies the key drivers of valence while excluding redundant variables like date to ensure a parsimonious model. The results demonstrate that danceability is the strongest predictor of valence, followed by tempo, acousticness, and scaled mean temperature, each showing statistically significant effects. These findings confirm the importance of both intrinsic song features and external contexts in shaping emotional perceptions of music.

This research contributes to the broader field of musicology and sentiment analysis by integrating environmental data with audio features, offering insights into the emotional dynamics of music consumption. The implications extend beyond academic interest, informing practical applications such as playlist curation, personalized recommendations, and mood-based marketing strategies. By enhancing our understanding of how music and context interact, this study provides a foundation for future explorations into the emotional landscape of modern music.

The remainder of this paper is structured as follows. Section 2 describes the dataset and measurement process, while Section 3 outlines the methodology and model validation. Section 4 presents the results, highlighting the key predictors of valence and their implications. Finally, Section 5 discusses the findings, limitations, and potential avenues for future research.

Liu, J., & Tzanetakis, G. (2020). Music Mood Classification: A Survey. IEEE Transactions on Multimedia.

Spotify for Developers. (2024). Spotify Charts API Documentation.

@@

## Overview {#sec-overview}

To analyze the relationship between song characteristics, temperature, and valence---which represents the mood or positivity of a song [@cite]---data from Spotify and weather records from Canada's most populous cities were integrated. Daily top 50 songs in Canada from Spotify Charts [@cite] were collected for the period between June 1, 2024, and November 16, 2024. Using the Spotify API, features such as Valence, Danceability, Acousticness, and Tempo were retrieved. Concurrently, weather data was obtained from the Government of Canada's Climate Weather website [@cite] for six major cities: Toronto, Montreal, Edmonton, Calgary, Ottawa, and Vancouver.

Data cleaning was conducted in R [@cite] using the tidyverse [@cite] and arrow [@cite] packages. The cities mentioned were selected due to their large populations, and a new numerical variable, Mean Temp, was created by weighting each city's daily temperature by its population density and dividing by the total population density. The resulting dataset consists of 8,450 observations and includes key variables such as Valence, Danceability, Acousticness, Tempo, Temperature, and Artist. The dataset was confirmed to have no missing or incomplete values, ensuring its reliability for analysis. A sample of the cleaned dataset is displayed in [@tbl-spotifysongs].

For visualization and analysis, the knitr [@cite], ggplot2 [@cite], and kableExtra [@cite] packages were utilized to generate professional tables and visualizations, effectively enhancing the presentation of results.

```{r}
#| label: tbl-spotifysongs
#| tbl-cap: Sample of Song's Characteristics
#| echo: false
#| warning: false
#| fig-align: center
analysis_data |> 
  select(Artist, Song, Valence, `Mean Temp`, Danceability, Acousticness, Tempo, Date) |> 
  mutate(
    `Mean Temp` = round(`Mean Temp`, 2),
    Acousticness = round(Acousticness, 3),
    Date = format(as.Date(Date), "%b %d") 
  ) |> 
  head(5) |> 
  kable(
    col.names = c("Artist", "Song", "Valence", "Mean Temp", "Dance", "Acoustic", "Tempo", "Date"),
  ) |> 
  kable_styling(
    full_width = FALSE, 
    position = "center", 
    font_size = 10
  ) |> 
  column_spec(1, width = "2.5cm") |>  # Adjust Artist column width
  column_spec(2, width = "2.7cm") |>  # Adjust Song column width
  column_spec(3:8, width = "1.25cm") 
```

[@tbl-summary] presents the summary statistics of the numerical variables reveal key insights into the dataset. Valence, which ranges from 0 to 1, spans 0.036 to 0.981, with a mean of 0.532, reflecting a wide spectrum of emotional tones from very low to highly positive tracks\[perhaps moderate mood\]. Mean Temperature, calculated as a population-weighted average across major Canadian cities, has a mean of 16.9°C, with values spanning from 3.5°C to 24.8°C, showcasing the climatic variation during the study period.

Danceability and Acousticness, both measured on a 0 to 1 scale, highlight contrasting patterns. Danceability values range from 0.234 to 0.943, with a mean of 0.632, indicating that most songs are moderately to highly danceable. Acousticness, ranging from 0.000 to 0.968, has a lower mean of 0.228, suggesting that non-acoustic tracks---those with more electronic or synthetic instrumentation---dominate the dataset.

Finally, Tempo, measured in beats per minute (BPM), has a mean of 126 BPM, with a wide range from 48.7 BPM for slower tracks to 203.8 BPM for faster ones. This diversity in tempo underscores the variety of musical styles included in the dataset, forming a robust basis for exploring their relationship with valence.

```{r}
#| label: tbl-summary
#| tbl-cap: Summary Statistics
#| echo: false
#| warning: false
#| fig-align: center
summary_table <- analysis_data %>%
  reframe(
    Variable = c("Valence", "Mean Temp", "Danceability", "Acousticness", "Tempo"),
    Mean = c(
      mean(Valence, na.rm = TRUE),
      mean(`Mean Temp`, na.rm = TRUE),
      mean(Danceability, na.rm = TRUE),
      mean(Acousticness, na.rm = TRUE),
      mean(Tempo, na.rm = TRUE)
    ),
    SD = c(
      sd(Valence, na.rm = TRUE),
      sd(`Mean Temp`, na.rm = TRUE),
      sd(Danceability, na.rm = TRUE),
      sd(Acousticness, na.rm = TRUE),
      sd(Tempo, na.rm = TRUE)
    ),
    Min = c(
      min(Valence, na.rm = TRUE),
      min(`Mean Temp`, na.rm = TRUE),
      min(Danceability, na.rm = TRUE),
      min(Acousticness, na.rm = TRUE),
      min(Tempo, na.rm = TRUE)
    ),
    Max = c(
      max(Valence, na.rm = TRUE),
      max(`Mean Temp`, na.rm = TRUE),
      max(Danceability, na.rm = TRUE),
      max(Acousticness, na.rm = TRUE),
      max(Tempo, na.rm = TRUE)
    )
  ) %>%
  mutate(
    Mean = round(Mean, 3),
    SD = round(SD, 3),
    Min = round(Min, 3),
    Max = round(Max, 3)
  )
kable(summary_table, format = "pipe")
```

The weather data used in this study originates from the Government of Canada's Climate Weather website [@cite], which collects meteorological observations from a comprehensive network of weather stations across the country. These stations measure key variables such as temperature, precipitation, and wind speed using standardized instruments and methodologies. Data is captured hourly or daily, depending on the station, and is validated through automated quality control processes and manual verification by Environment and Climate Change Canada. This rigorous approach ensures reliable and precise measurements that meet international meteorological standards. In the context of this analysis, one weather station per city was selected based on relevance and proximity; additional details on this selection process are provided in the appendix section.

The music data originates from Spotify Charts [@cite], which rank the most-streamed songs in a given region daily. Spotify compiles these rankings by aggregating anonymized user streaming data across its platform. Streams are verified to ensure authenticity, eliminating artificial plays generated by bots or fraudulent activity. In addition to the rankings, Spotify employs machine learning models to analyze the audio features of each song, such as Valence, Danceability, Acousticness, and Tempo. These features are derived from the acoustic and digital signal processing of tracks, providing a quantitative representation of each song's characteristics.

The combination of these trusted data sources ensures that the dataset used in this study is grounded in accurate and validated measurements. Weather data reflects real-world environmental conditions, while Spotify's data captures both user engagement and precise audio analysis, enabling robust exploration of the relationship between temperature, song features, and valence.

## Analysis of Variables

### Outcome variable

The Valence variable exhibits a fairly uniform distribution, as shown in [@fig-valence], with values ranging from 0.036 to 0.981, as detailed in [@tbl-summary]. This wide range captures a broad spectrum of emotional tones in the dataset.

```{r}
#| label: fig-valence
#| fig-cap: Distribution of Valence
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4

ggplot(analysis_data, aes(x = Valence)) +
  geom_histogram(binwidth = 0.05, fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(x = "Valence", y = "Count") +
  theme_minimal()
```

[@tbl-highest] highlights the top 5 unique songs with the highest Valence, led by "September" by Earth, Wind & Fire (0.981), alongside other highly positive tracks like "Apple" by Charli XCX (0.962) and "Super Freak" by Rick James (0.962). These songs represent the most uplifting and positive moods in the dataset.

```{r}
#| label: tbl-highest
#| tbl-cap: Top 5 Songs with the Highest Valence
#| echo: false
#| warning: false
#| tbl-align: center
# Show the top 5 songs with the highest Valence
top_valence_songs_unique <- analysis_data %>%
  distinct(Song, Artist, Valence) %>%
  arrange(desc(Valence)) %>%
  head(5)

kable(top_valence_songs_unique, col.names = c("Song", "Artist", "Valence"))
```

In contrast, [@tbl-lowest] features the top 5 unique songs with the lowest Valence, including "BLUE" by Billie Eilish (0.0365) and "Meteor Man" by Lil Uzi Vert (0.0384), which reflect darker or more somber tones. Together, these extremes illustrate the dataset's diversity, spanning a wide emotional range that provides a solid foundation for analyzing relationships with song features and external factors.

```{r}
#| label: tbl-lowest
#| tbl-cap: Top 5 Songs with the Lowest Valence
#| echo: false
#| warning: false
#| tbl-align: center
# Show the top 5 unique songs with the lowest Valence
lowest_valence_songs_unique <- analysis_data %>%
  distinct(Song, Artist, Valence) %>% 
  arrange(Valence) %>%
  head(5)

kable(lowest_valence_songs_unique, col.names = c("Song", "Artist", "Valence"))
```

### Predictor variables

-   **Mean Temperature**

[@fig-temp] shows the average temperature over time, demonstrating a seasonal decline from approximately 25°C in the summer to below 5°C in late autumn. This pattern aligns with expected seasonal transitions in Canada and is further supported by [@tbl-summary], where the Mean Temperature averages 16.929°C with a standard deviation of 5.298°C. This notable variability offers an opportunity to explore potential correlations between external environmental factors, like weather, and shifts in musical preferences and emotional positivity as reflected in Valence.

```{r}
#| label: fig-temp
#| fig-cap: Average Temperature from June to November 2024
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4

ggplot(analysis_data, aes(x = Date, y = `Mean Temp`)) +
  geom_line(color = "darkblue", alpha = 0.7) +
  labs(x = "Date", y = "Temperature (°C)") +
  theme_minimal()
```

-   **Artist**

The variable Artist reveals intriguing patterns in the dataset. Displayed in [@fig-artist], the top 10 artists, led by Sabrina Carpenter and Zach Bryan, contributed the highest number of tracks to the dataset, each exceeding 600 tracks, underscoring their strong presence on the charts during the study period. However, despite their prominence, their songs do not appear in [@tbl-highest] or [@tbl-lowest], unlike artists such as Chappell Roan and Billie Eilish, whose tracks span the emotional spectrum by featuring in both the highest and lowest Valence categories. Hozier, contributing the fewest tracks among the top 10, stands out for appearing in [@tbl-highest], showcasing the emotional positivity of some of his music despite a smaller overall presence.

```{r}
#| label: fig-artist
#| fig-cap: Top 10 Artists by Number of Tracks
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4
# Get the top 10 artists by song count
top_artists <- analysis_data %>%
  count(Artist, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  mutate(Artist = gsub(" ", "\n", Artist))

ggplot(top_artists, aes(x = reorder(Artist, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
  labs(x = "Artist",
       y = "Number of Tracks") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0.5, size = 6))
```

-   **Tempo**

[@fig-tempo] illustrates the distribution of Tempo, revealing a roughly normal shape. The values predominantly cluster between 100 and 130 BPM, with the highest frequency near 120 BPM, reflecting the common tempo range of popular music. This pattern suggests a tendency for moderate-paced tracks to dominate the dataset, which aligns with the rhythmic preferences in mainstream music consumption.

In addition, the distribution highlights outliers on both ends of the spectrum, with slower tempos below 60 BPM and faster tempos above 180 BPM representing niche genres or unique musical styles.

```{r}
#| label: fig-tempo
#| fig-cap: Distribution of Tempo
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4

ggplot(analysis_data, aes(x = Tempo)) +
  geom_histogram(binwidth = 10, fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(x = "Tempo (BPM)", y = "Frequency") +
  theme_minimal()
```

-   **Danceability and Acousticness**

The [@fig-diff] below highlights distinct patterns in the distributions of Danceability and Acousticness.

Acousticness displays a strongly right-skewed distribution, with most values concentrated near 0, reaffirming the prevalence of non-acoustic tracks in the dataset as noted in [@sec-overview]. This pattern aligns with its low mean of 0.228 and relatively high variability, as shown in [@tbl-summary], capturing a mix of purely electronic tracks and a smaller presence of acoustic elements.

In contrast, Danceability demonstrates a more uniform distribution, peaking between 0.6 and 0.8. With a higher mean of 0.632 and a standard deviation of 0.134, this indicates that the majority of tracks are moderately to highly danceable. This distribution underscores the dataset's emphasis on popular music, where rhythmic appeal and danceability are often prioritized.

```{r}
#| label: fig-diff
#| fig-cap: Distribution of Danceability and Acousticness
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 7
#| fig.height: 5

analysis_data %>%
  select(Danceability, Acousticness) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value, fill = Variable)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7, position = "identity", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  scale_fill_manual(values = c("Danceability" = "#4B0082", "Acousticness" = "#008080")) + # Custom colors
  labs(
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Exploring Relationships

-   **Valence vs. Temperature**

The heatmap in [@fig-heatmap] highlights that higher temperatures (20-25°C) are associated with a higher frequency of moderately positive tracks, with Valence values clustering around 0.5-0.75. In contrast, lower temperatures (5-15°C) show a similar distribution, with song positivity largely centered in the mid-range, indicating no significant skew toward extreme positivity or negativity. The lack of data in the lowest temperature range (0-5°C) reflects its absence in the dataset, rather than a meaningful trend.

```{r}
#| label: fig-heatmap
#| fig-cap: "Heatmap of Valence vs. Temperature"
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4

analysis_data %>%
  mutate(Temp_Range = cut(`Mean Temp`, breaks = seq(0, 25, by = 5))) %>%
  ggplot(aes(x = Temp_Range, y = Valence, fill = after_stat(count))) + 
  stat_bin2d(binwidth = c(1, 0.1)) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
  labs(x = "Temperature Range (°C)",
    y = "Valence (Happiness)"
  ) +
  theme_minimal() +
  annotate(
    "text", x = 1, y = 0.05, label = "No data in this region", color = "darkred", size = 2.5
  ) +
  theme (
    legend.key.size = unit(0.4, "cm"), 
    legend.text = element_text(size = 7), 
    legend.title = element_text(size = 8) 
  )
```

-   **Valence vs. Artist**

The chart in [@fig-avgvalence] highlights notable contrasts in the emotional tone of music produced by the top 10 artists. Hozier, Tommy Richman, and Chappell Roan stand out with the highest average Valence scores, suggesting their tracks lean toward more positive and uplifting moods. This trend aligns with their reputations for creating emotionally resonant music. Conversely, Kendrick Lamar and Zach Bryan, who rank lowest in average Valence, demonstrate a preference for deeper or more introspective themes, reflecting their storytelling-driven styles. The diversity in Valence among these artists underscores the variety of emotional experiences their music offers, capturing a broad spectrum of moods within the dataset.

```{r}
#| label: fig-avgvalence
#| fig-cap: Average Valence by Top 10 Artists
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4

# Average Valence by Artist (Top 10 Artists)
top_artists <- analysis_data %>%
  group_by(Artist) %>%
  summarize(avg_valence = mean(Valence), count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  mutate(Artist = gsub(" ", "\n", Artist))

ggplot(top_artists, aes(x = reorder(Artist, avg_valence), y = avg_valence, fill = avg_valence)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Artist",
    y = "Average Valence"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(hjust = 0.5, size = 6))
```

@@

This section presents a multiple linear regression (MLR) model designed to predict the valence of a song, reflecting its perceived positivity, based on audio features and contextual factors. The model assesses how characteristics such as danceability, acousticness, scaled tempo, and scaled mean temperature, as well as artist-specific effects, contribute to the song's valence. The model provides a straightforward and interpretable framework for analyzing the linear relationships between predictors and song valence.

To conduct this analysis, the R packages lm [@cite] and MLmetrics [@cite] were employed. The lm function facilitates the specification of the regression model, while the MLmetrics package was used to calculate performance metrics such as Root Mean Square Error (RMSE) and R-squared for model evaluation.

## Model Structure

$$
\begin{align}
y_i &= \beta_0 + \beta_1 (\text{Scaled Mean Temp})_i + \beta_2 (\text{Artist})_i + \beta_3 (\text{Danceability})_i \\
    &\quad + \beta_4 (\text{Acousticness})_i + \beta_5 (\text{Scaled Tempo})_i + \epsilon_i \\
\end{align}
$$

Where:

$$
\begin{align}
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}
$$

-   $y_i$: Valence of the song, measuring perceived positivity.
-   $\beta_0$: Baseline valence when predictors are at their reference levels.
-   $\beta_1$: Effect of scaled mean temperature on valence.
-   $\beta_2$: Artist-specific effects on valence.
-   $\beta_3$: Effect of danceability on valence.
-   $\beta_4$: Effect of acousticness on valence.
-   $\beta_5$: Effect of scaled tempo on valence.
-   $\epsilon_i$: Unexplained variability, assumed to follow a normal distribution with mean 0 and variance.

## Variable Selection

In constructing the MLR model, careful consideration was given to selecting predictors that accurately capture the relationships influencing song valence:

-   Scaled Mean Temperature: Captures the contextual influence of environmental factors on emotional responses to music. Scaling was applied to normalize its range and enhance interpretability alongside other predictors.

-   Artist: Accounts for the unique contributions of individual artists to song valence, capturing categorical variability in positivity perception.

-   Danceability: Measures the rhythmic and energetic qualities of a song, which are strongly associated with listener enjoyment and positive emotional responses.

-   Acousticness: Reflects the extent to which a track is perceived as acoustic, influencing the mood and perceived positivity of the song.

-   Scaled Tempo: Tempo impacts energy and mood. Scaling ensures comparability across its wide range, allowing for a consistent assessment of its effects on valence.

These variables were selected based on their theoretical relevance and potential to explain variations in song positivity while ensuring robust model performance.

## Model Validation & Justification

- **Model Validation**

The model validation process involved out-of-sample testing, where the dataset was split into training and testing sets in a 70-30 ratio. Three variations of the multiple linear regression (MLR) model were tested, differing in the number of predictors included. Model performance was evaluated using Root Mean Square Error (RMSE) and R-squared values. As shown in [@tbl-comp], the models demonstrate varying degrees of predictive accuracy and explanatory power.
```{r}
#| label: tbl-comp
#| fig-cap: Comparison of Model Performance
#| echo: false
#| warning: false
#| fig-align: center

comparison_data <- data.frame(
  Model = c("MLR 1", "MLR 2", "MLR 3"),
  Variables = c(
    "Scaled Mean Temp, Artist, Danceability,<br>Acousticness, Scaled Tempo, Date",
    "Scaled Mean Temp, Artist, Danceability,<br>Acousticness, Scaled Tempo",
    "Scaled Mean Temp, Danceability,<br>Acousticness, Scaled Tempo"
  ),
  RMSE = c(0.0868, 0.0866, 0.2031),
  `R-Squared` = c(0.8597, 0.8601, -1.4267),
  check.names = FALSE
)

knitr::kable(
  comparison_data,
  format = "pipe",
  align = "l",
  col.names = c("Model", "Variables", "R-Squared", "RSME")
)
```
After not taking into consideration Artist, MLR 3 includes fewer predictors, it performs significantly worse, as evidenced by its negative R-squared and much higher RMSE, indicating poor fit and predictive ability. On the other hand, MLR 1 and MLR 2 exhibit comparable performance, with MLR 2 slightly outperforming MLR 1 in terms of both RMSE and R-squared.

- **Model Justification**
The choice of MLR 2 is justified by its balance of predictive performance and model simplicity. The `Date` variable was excluded because it is indirectly related to `Scaled Mean Temp`, which already captures the temporal context in a more meaningful way. Including `Date` could introduce redundancy and unnecessary complexity, detracting from the model's parsimony. Furthermore, this model includes the key variables of interest outlined in the previous section, ensuring it addresses the research objectives effectively. Additionally, MLR 2 slightly outperforms MLR 1 in terms of RMSE and R-squared, further supporting its selection as the final model for predicting song valence.

- **Diagnostics**
A plot comparing the predicted and true valence values was generated to visually assess the model's performance [@fig-predicted]. The dashed line represents the 1:1 relationship, indicating perfect predictions. Most points cluster around the line, showing that the model effectively captures the relationship between predictors and valence. However, some deviations are observed, which may indicate areas where the model could be improved.

@@

The results from the multiple linear regression (MLR) analysis reveal significant insights into the predictors of valence, as shown in the coefficient summary (Table 1) and the corresponding coefficient plot (Figure 1). These tables highlight the statistical significance of the predictors, measured by their respective t-statistics and standard errors, which confirm the reliability of the estimates.

Danceability emerges as the most impactful and statistically significant predictor, with a large positive coefficient (1.260) and a t-statistic of 65.998, far exceeding the critical threshold for significance. This underscores that higher danceability strongly correlates with increased valence, a result consistent with the idea that rhythmic and energetic qualities in music enhance perceived positivity.

Scaled Tempo also demonstrates a statistically significant positive effect on valence, with a coefficient of 0.059 and a t-statistic of 32.527. Its standardized impact highlights the relationship between faster-paced songs and heightened positivity. Similarly, Scaled Mean Temperature, while having a smaller coefficient of 0.007, remains statistically significant with a t-statistic of 4.351, reflecting its nuanced role in influencing music perception in relation to environmental context.

In contrast, Acousticness exhibits a significant negative effect on valence, with a coefficient of -0.261 and a t-statistic of -26.327. This result suggests that tracks with strong acoustic features, often associated with somber or introspective tones, are generally perceived as less positive. The inclusion of this variable provides valuable insights into how specific audio characteristics influence the emotional reception of songs.

The categorical variable Artist captures substantial variability in song valence, as evidenced by the statistically significant coefficients for specific artists. For instance, Brenda Lee and Dean Martin positively influence valence, with coefficients of 0.403 and 0.351, respectively, and t-statistics exceeding 4. Notably, artists such as Eminem (-0.997, t = -28.162) and Kendrick Lamar (-0.873, t = -47.717) exert a negative influence on valence, reflecting stylistic differences and emotional themes in their music. These results are highlighted by selecting the most impactful artists in Table 1 to focus on those with substantial effects.

The Date variable, while included in the analysis, shows minimal relevance with a coefficient near zero (0.000) and a low t-statistic of 3.906, further supporting its exclusion in the final model (MLR 2). This confirms that the temporal aspect of the dataset adds little explanatory value beyond other predictors such as Scaled Mean Temperature.

The coefficient plot (Figure 1) visually reinforces these findings, illustrating the magnitude and statistical significance of each predictor's effect. Predictors with larger t-statistics and narrower confidence intervals demonstrate greater reliability in their estimated effects, aligning with theoretical expectations.

Overall, the statistical significance of the coefficients validates the robustness of the MLR model. Key predictors such as Danceability, Tempo, Acousticness, and Scaled Mean Temperature exhibit strong and reliable associations with valence, while the inclusion of impactful artists provides a nuanced understanding of categorical variability. These findings underscore the model’s ability to capture the intricate relationships between audio features, artist effects, and song positivity.

```{r}
#| echo: false
#| eval: true
#| warning: false
mlm_data_analysis <- analysis_data %>%
  mutate(predicted_valence_lm = predict(mlm_model_analysis, newdata = analysis_data))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: fsk
#| warning: false

# Extract coefficients from the model
tidy_model <- tidy(mlm_model)

# Filter coefficients related to Artist
artist_coefficients <- tidy_model %>%
  filter(str_detect(term, "Artist")) %>%
  arrange(desc(abs(estimate)))

# Top 5 most positive and negative impacts
top_positive <- artist_coefficients %>%
  slice_max(order_by = estimate, n = 5)

top_negative <- artist_coefficients %>%
  slice_min(order_by = estimate, n = 5)

# Combine results for impactful artists
impactful_artists <- bind_rows(top_positive, top_negative)

# Add non-artist terms back for a complete table
filtered_model <- tidy_model %>%
  filter(term %in% impactful_artists$term | !str_detect(term, "Artist"))

# Render a table with kableExtra
filtered_model %>%
  mutate(term = str_replace(term, "Artist", "")) %>%
  select(term, estimate, std.error, statistic) %>%
  kable(
    col.names = c("Term", "Estimate", "Std. Error", "Statistic"),
    format = "html",
    digits = 3
  ) %>%
  kable_styling(full_width = FALSE, position = "center")
```

```{r}
# Extract coefficients from the model
tidy_model <- broom::tidy(mlm_model)

# Filter coefficients related to Artist
artist_coefficients <- tidy_model %>%
  filter(str_detect(term, "Artist")) %>%
  arrange(desc(abs(estimate)))

# Top 5 most positive and negative impacts
top_positive <- artist_coefficients %>%
  slice_max(order_by = estimate, n = 5)

top_negative <- artist_coefficients %>%
  slice_min(order_by = estimate, n = 5)

# Combine results for impactful artists
impactful_artists <- bind_rows(top_positive, top_negative)

# Add non-artist terms back for a complete dataset
filtered_model <- tidy_model %>%
  filter(term %in% impactful_artists$term | !str_detect(term, "Artist"))

# Plot the filtered coefficients
ggplot(filtered_model, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(color = "blue", size = 3) +  # Points for coefficients
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error, ymax = estimate + 1.96 * std.error), 
                width = 0.2, color = "gray") +  # Error bars for confidence intervals
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Top Artist Impacts and Other Predictors",
    x = "Predictors",
    y = "Coefficient Estimate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text = element_text(color = "gray20"),
    axis.title = element_text(color = "gray30")
  )
```

@@

**What Was Done in This Paper**

This paper employs a multiple linear regression (MLR) framework to investigate the relationships between audio features, contextual variables, and song valence. By integrating Spotify's audio feature data with contextual information such as population-weighted mean temperature, this study identifies and quantifies the contributions of key predictors to valence. Predictors such as danceability, acousticness, and tempo were examined alongside artist-specific effects to evaluate their respective roles in shaping the emotional tone of music. The analysis highlights statistically significant effects for most predictors, confirming their relevance in predicting valence, while excluding redundant variables like date to ensure a parsimonious model.

**Insights Gained About the World**

The results underline the critical role of intrinsic audio features in determining a song's valence. Danceability emerges as the most influential factor, suggesting that rhythmic and energetic qualities strongly resonate with listeners' perceptions of positivity. Similarly, tempo and acousticness demonstrate statistically significant effects, with faster tempos generally associated with more positive emotional tones and acoustic tracks correlating with more introspective or somber moods. These findings affirm the connection between specific musical attributes and listener sentiment, contributing to our understanding of the emotional dynamics of music.

Another key insight relates to the contextual influence of mean temperature. Although its effect is smaller compared to audio features, the statistically significant relationship between temperature and valence suggests that environmental factors subtly shape how listeners perceive the emotional tone of music. This highlights the potential for integrating external contextual data into music analysis, offering novel avenues for exploring how environment and culture influence musical preferences.

**Weaknesses of the Approach**
